---
title: "Figure 2A — niche summary"
author: "BCRL Multi-Omics"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    fig_width: 15
params:
  results_dir: "results"
  fdr: 0.05
  # --- prevalence ---------------------------------------------------------------
  min_hex: 40           # hexbins a niche needs in a section before it counts as present
  block_width: 0.52     # prevalence block width (1 = blocks touch)
  block_height: 0.6
  # --- density glow -------------------------------------------------------------
  ball_min: 2.4         # ring diameter (mm) for the least dense niche
  ball_max: 4.4         # ...and for the densest
  glow_size: 1.0        # mm ADDED around the ring -- not a multiplier. > ~1.2 and
                        # neighbouring halos touch at the current figure height
  glow_alpha: 0.30
  # --- pathway dots -------------------------------------------------------------
  dot_ns: 2.8           # dot diameter for n.s.
  dot_sig: 6.5          # ...and for FDR < fdr
  # --- layout -------------------------------------------------------------------
  fig_width: 15
  fig_height: 6.0
  widths: !r c(0.62, 0.75, 0.6, 1.0, 1.15, 3.0)   # niche, abundance, density, prevalence, program, dots
  save: TRUE            # FALSE while tuning: preview only, nothing written to results/
---

Tuning copy of `fig2a_composite.R`. Same inputs and same numbers — every knob I would
otherwise edit in the script is a `param` above, and each panel prints on its own so you can
re-knit one chunk at a time. Set `save: FALSE` to preview without overwriting `results/Fig2A.*`.

Built to `Fig2A_concept.svg`: niche swatch, abundance bar, density glow balls, prevalence
blocks, transcriptional program, and the ROS + signalling scores as dots.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, dev = "png", dpi = 150)
suppressMessages({
  library(ggplot2); library(dplyr); library(readr); library(forcats); library(tidyr)
  library(patchwork)
})
RES <- params$results_dir
FDR <- params$fdr
# hexbin object behind the old prevalence donuts (scripts/figures/niche_prev_density_code.r)
SF_HEX  <- "/vast/projects/BCRL_Multi_Omics/scripts/figures/sf_hex.rds"
# cell-level object behind the old density panel; only read when the density cache is absent
VEN_ALL <- "/vast/projects/BCRL_Multi_Omics/scripts/figures/ven_all.rds"
```

## Palette, order and labels

`NORDER` sets the row order top to bottom — reorder it and every panel follows. `PROGRAM` is
transcribed from the concept SVG and is **not** in any results CSV, so check it against the
semantic modules before publishing.

```{r constants}
REN <- c(T_A="T-ROS", T_TC="T-PAN", T_P="T-P", T_AA="T-AC", T_AMN="T-MES", T_LE="T-OC",
         T_OM="T-GL", T_O="T-MET2", T_M="T-MET1", I="I", V="V", N="N")
NICHE_COLORS_OLD <- c(T_LE="#9FCF87", T_AMN="#94FFB5", T_P="#9068AC", V="#F77800",
                      T_A="#FFCC99", T_M="#2BCE48", I="#C39C94", N="#005C31",
                      T_TC="#0075DC", T_O="#F0A0FF", T_OM="#C20088", T_AA="#4C005C")
NCOL <- setNames(unname(NICHE_COLORS_OLD), unname(REN[names(NICHE_COLORS_OLD)]))

NORDER <- c("T-AC","T-PAN","T-MES","T-MET2","T-MET1","T-P","T-ROS","T-GL")

PROGRAM <- c("T-ROS"  = "mRNA processing, chromatin",
             "T-MES"  = "Cilium assembly, SUMOylation",
             "T-P"    = "Mitosis, G2M checkpoint",
             "T-GL"   = "mRNA splicing, chromatin",
             "T-AC"   = "Cholesterol, gap junctions",
             "T-PAN"  = "Synaptic transmission",
             "T-MET1" = "Synaptic vesicle clustering",
             "T-MET2" = "Catecholamine, serotonin uptake")

theme_set(theme_minimal(base_size = 11))
# the columns share one y axis; only the first panel prints the niche names
bare <- theme(panel.grid = element_blank(),
              axis.title = element_blank(),
              axis.text.y = element_blank(),
              axis.ticks = element_blank(),
              plot.title = element_text(size = 10, hjust = 0, face = "plain"),
              plot.margin = margin(4, 6, 4, 6))
```

## Abundance — mean share of a specimen

Absences count as 0, so a niche confined to a few sections does not outrank one found
everywhere. Averaging only over the samples where a niche occurs puts T-MES top instead.

```{r composition}
cd <- read_csv(file.path(RES, "aggregated_filtered_coldata.csv"), show_col_types = FALSE) %>%
  transmute(patient, niche = gsub("_", "-", niche), ncells) %>%
  group_by(patient, niche) %>% summarise(ncells = sum(ncells), .groups = "drop")

N_SPEC <- n_distinct(cd$patient)

comp <- cd %>%
  group_by(patient) %>% mutate(frac = ncells / sum(ncells)) %>% ungroup() %>%
  complete(patient, niche, fill = list(ncells = 0, frac = 0)) %>%
  filter(niche %in% NORDER) %>%
  group_by(niche) %>%
  summarise(abundance = mean(frac), n_spec = sum(ncells > 0), .groups = "drop") %>%
  mutate(nn = factor(niche, levels = NORDER))
comp %>% arrange(desc(abundance)) %>% mutate(abundance = sprintf("%.1f%%", 100 * abundance))
```

## Prevalence — the old donut figure's rule

`colSums(table(section, niche) >= min_hex) / n_sections`, exactly as
`scripts/figures/niche_prev_density_code.r` scored it: a niche counts for a section only once
it holds at least `params$min_hex` hexbins there. The 26 ids are **sections**, not patients —
collapsing them to 15 patients is one `rowsum()` away if you ever want one vote per patient.

```{r prevalence}
MIN_HEX <- params$min_hex
NUM2OLD <- c("0"="T_LE","1"="T_AMN","2"="T_P","3"="V","4"="T_A","5"="T_M",
             "6"="I","7"="N","8"="T_TC","9"="T_O","10"="T_OM","11"="T_AA")

sf_hex <- readRDS(SF_HEX)
tb     <- table(sf_hex$patient, sf_hex$niche)
N_PAT  <- nrow(tb)
hits   <- colSums(tb >= MIN_HEX)

prev <- tibble(cluster = names(hits), n_pat = as.integer(hits)) %>%
  mutate(niche = unname(REN[NUM2OLD[cluster]]),   # cluster id -> old code -> new name
         prevalence = n_pat / N_PAT) %>%
  filter(!is.na(niche), niche %in% NORDER) %>%    # drops the "white" background class
  mutate(nn = factor(niche, levels = NORDER))
stopifnot(nrow(prev) == length(NORDER))
prev %>% arrange(desc(prevalence)) %>% transmute(niche, n = sprintf("%d/%d", n_pat, N_PAT))
```

## Density — mean cells per hexbin

The other half of the old script: each hexbin goes to the niche holding most of its cells,
density is the mean cell count of that niche's hexbins. Needs `ven_all.rds` (1.1 GB, ~60 s),
so it is cached to `results/niche_hex_density.csv` — delete that file to recompute.

```{r density-data}
DENS_CSV <- file.path(RES, "niche_hex_density.csv")
if (!file.exists(DENS_CSV)) {
  message("computing hexbin density from ", VEN_ALL, " (this loads 1.1 GB)")
  ven <- readRDS(VEN_ALL)
  htb <- table(ven$hexbin_id, ven$niche)
  hex_df <- data.frame(count = rowSums(htb), niche = colnames(htb)[max.col(htb)])
  d <- vapply(split(hex_df$count, hex_df$niche), mean, numeric(1))
  write_csv(tibble(cluster = names(d), density = as.numeric(d),
                   n_hexbins = as.integer(table(hex_df$niche)[names(d)])), DENS_CSV)
  rm(ven, htb); gc()
}

dens <- read_csv(DENS_CSV, show_col_types = FALSE) %>%
  # ven_all labels niches with the OLD hyphenated codes (T-A, T-AA, ...), not cluster ids
  mutate(niche = unname(REN[gsub("-", "_", cluster)])) %>%
  filter(!is.na(niche), niche %in% NORDER) %>%
  mutate(nn = factor(niche, levels = NORDER))
stopifnot(nrow(dens) == length(NORDER))
dens %>% arrange(desc(density)) %>% transmute(niche, density = round(density, 1))
```

# Panels

Each chunk below builds one column and prints it, so you can re-run just the one you are
tuning. The assembly chunk at the end puts them together.

## 1. Niche swatch

```{r panel-name, fig.width=2, fig.height=3.2}
p_name <- ggplot(comp, aes(0, fct_rev(nn))) +
  geom_tile(aes(fill = nn), width = 0.5, height = 0.5) +
  geom_text(aes(x = 0.45, label = nn), hjust = 0, size = 3.5) +
  scale_fill_manual(values = NCOL, guide = "none") +
  scale_x_continuous(limits = c(-0.35, 2.4), expand = c(0, 0)) +
  labs(title = "Niche") +
  bare + theme(axis.text.x = element_blank())
p_name
```

## 2. Abundance bar

```{r panel-abundance, fig.width=3, fig.height=3.2}
p_ab <- ggplot(comp, aes(abundance, fct_rev(nn), fill = nn)) +
  geom_col(width = 0.62) +
  scale_fill_manual(values = NCOL, guide = "none") +
  scale_x_continuous(breaks = c(0, max(comp$abundance)), labels = c("0", "max"),
                     expand = expansion(mult = c(0, 0.04))) +
  labs(title = "Abundance", subtitle = "mean share of a specimen") +
  bare + theme(axis.text.x = element_text(size = 8, colour = "grey40"),
               plot.subtitle = element_text(size = 8, colour = "grey40"))
p_ab
```

## 3. Density glow balls

An unfilled grey ring with a halo tinted by the niche colour — identity without a second
palette, size carrying the number.

**The knob that bites:** `glow_size` is millimetres *added* around the ring, not a multiplier.
The rows sit about `r sprintf("%.1f", 25.4 * params$fig_height * 0.55 / 8)` mm apart at the
current `fig_height`, so a ring of `ball_max` + 2 × `glow_size` beyond roughly that will make
neighbouring halos touch. Raise `fig_height` if you want both bigger.

Sizes are spread over the observed range (9.1–13.2), not from zero: area-proportional sizing
makes eight balls differing by <1.5× look identical. Read the column as a ranking, not a ratio.

```{r panel-density, fig.width=2, fig.height=3.2}
p_dn <- ggplot(dens, aes(1, fct_rev(nn))) +
  ggshadow::geom_glowpoint(aes(size = density, shadowcolour = nn), shape = 16,
                           colour = "grey50", alpha = 0.01,
                           shadowalpha = params$glow_alpha, shadowsize = params$glow_size) +
  ggshadow::scale_shadowcolour_manual(values = NCOL, guide = "none") +
  geom_point(aes(size = density), shape = 21, fill = "#FFFFFF03", colour = "grey35",
             stroke = 0.7) +
  scale_size(range = c(params$ball_min, params$ball_max), name = "mean cells\nper hexbin",
             breaks = c(9, 11, 13), limits = c(9, 13.3)) +
  scale_x_continuous(limits = c(0.55, 1.45), breaks = NULL, expand = c(0, 0)) +
  labs(title = "Density", subtitle = "mean cells per hexbin") +
  bare + theme(plot.subtitle = element_text(size = 8, colour = "grey40"),
               legend.position = "bottom", legend.title = element_text(size = 8),
               legend.text = element_text(size = 7.5), legend.key.size = unit(0.6, "lines"))
p_dn
```

## 4. Prevalence blocks

One block per section, filled where that niche clears the hexbin threshold. `block_width`
below 1 leaves the gaps; at 1 the blocks butt together into a bar.

```{r panel-prevalence, fig.width=4, fig.height=3.2}
pv_blocks <- prev %>%
  tidyr::expand(nesting(nn, n_pat), slot = 1:N_PAT) %>%
  mutate(on = slot <= n_pat)

p_pv <- ggplot(pv_blocks, aes(slot, fct_rev(nn))) +
  geom_tile(aes(fill = ifelse(on, as.character(nn), NA_character_)),
            colour = "grey60", linewidth = 0.2,
            width = params$block_width, height = params$block_height) +
  scale_fill_manual(values = NCOL, na.value = "white", guide = "none") +
  scale_x_continuous(breaks = NULL, expand = expansion(add = 0.5)) +
  labs(title = "Prevalence",
       subtitle = sprintf("patients with ≥%d hexbins (n/%d)", MIN_HEX, N_PAT)) +
  bare + theme(plot.subtitle = element_text(size = 8, colour = "grey40"))
p_pv
```

## 5. Transcriptional program

```{r panel-program, fig.width=4, fig.height=3.2}
p_pr <- ggplot(comp, aes(0, fct_rev(nn))) +
  geom_text(aes(label = PROGRAM[as.character(nn)]), hjust = 0, size = 3.3) +
  scale_x_continuous(limits = c(0, 1), expand = c(0, 0)) +
  labs(title = "Transcriptional program") +
  bare + theme(axis.text.x = element_blank())
p_pr
```

## 6. Pathway dot plot

Colour = mean logFC, size = significance in two steps, matching the concept's legend.

```{r panel-dots, fig.width=8, fig.height=4}
SORDER <- c("NOTCH","PDGF","RAS-MAPK","JAK-STAT","TGF-β","WNT","BMP")
ROSSET <- c("OXPHOS [GOBP]", "KEAP1-NFE2L2")

comb <- read_csv(file.path(RES, "combined_signalling_oxphos_nrf2.csv"), show_col_types = FALSE) %>%
  filter(niche %in% NORDER) %>%
  mutate(nn  = factor(niche, levels = NORDER),
         grp = factor(ifelse(set %in% ROSSET, "ROS", "Signalling pathways"),
                      levels = c("ROS", "Signalling pathways")),
         set = factor(set, levels = c(ROSSET, SORDER)),
         sig = FDR < !!FDR)
stopifnot(!any(is.na(comb$set)))

lim <- quantile(abs(comb$mean_logFC), 0.98, na.rm = TRUE)

p_dot <- ggplot(comb, aes(set, fct_rev(nn))) +
  geom_point(aes(fill = pmax(pmin(mean_logFC, lim), -lim), size = sig),
             shape = 21, colour = "grey45", stroke = 0.3) +
  facet_grid(. ~ grp, scales = "free_x", space = "free_x") +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0,
                       name = "Score", limits = c(-lim, lim),
                       # breaks pulled just inside the limits, else ggplot drops the end labels
                       breaks = c(-lim, lim) * 0.97, labels = c("low", "high"),
                       guide = guide_colourbar(display = "rectangles", barheight = 0.6,
                                               barwidth = 6, title.position = "top",
                                               direction = "horizontal")) +
  scale_size_manual(values = c(`FALSE` = params$dot_ns, `TRUE` = params$dot_sig),
                    labels = c(`FALSE` = "n.s.", `TRUE` = sprintf("FDR < %.2g", FDR)),
                    name = NULL) +
  guides(size = guide_legend(override.aes = list(fill = "white", colour = "grey45"))) +
  labs(x = NULL, y = NULL) +
  bare +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8.5),
        strip.text.x = element_text(size = 10),
        panel.grid.major = element_line(colour = "grey93"),
        legend.position = "bottom", legend.box = "horizontal",
        legend.title = element_text(size = 9), legend.text = element_text(size = 8))
p_dot
```

# Assembled figure

`params$widths` is the relative width of each column, in the order they appear. patchwork
aligns the panel areas, so the rows line up across all six regardless of what the axes below
them take up.

```{r assemble, fig.width=params$fig_width, fig.height=params$fig_height}
fig <- p_name + p_ab + p_dn + p_pv + p_pr + p_dot +
  plot_layout(nrow = 1, widths = params$widths) +
  plot_annotation(title = "Figure 2A — niche summary",
                  theme = theme(plot.title = element_text(face = "bold", size = 13)))
fig
```

```{r save, eval=params$save}
# svglite, never cairo's svg(): cairo composites through a filter + luminance masks that
# Affinity Publisher ignores, importing the whole figure as solid black.
ggsave(file.path(RES, "Fig2A.svg"), fig, device = svglite::svglite,
       width = params$fig_width, height = params$fig_height)
ggsave(file.path(RES, "Fig2A.png"), fig,
       width = params$fig_width, height = params$fig_height, dpi = 200)
cat("wrote", file.path(RES, c("Fig2A.svg", "Fig2A.png")), sep = "\n  ")
```

## The numbers behind the figure

```{r summary}
comp %>% left_join(prev, by = "niche") %>% left_join(dens, by = "niche") %>%
  arrange(match(niche, NORDER)) %>%
  transmute(niche, abundance = sprintf("%.1f%%", 100 * abundance),
            density = round(density, 1),
            prevalence = sprintf("%d/%d (%.0f%%)", n_pat, N_PAT, 100 * prevalence))
```
